---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "Amanda Sinofsky"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

There is a much higher percentage of FARMS in Moco MD than in Moco PA. However, the math scores percentages are not too different between the two even though there is a big difference in FARMS. This concerns me and makes me think that the two variables are not that related. 


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

This scatterplot is kind of all over the place, but it shows a negative relationship between FARMS percentage and math scores, indicating that the higher the farms percentage is the lower the math scores are. They might look so different because household income might be lower in Moco MD than in Moco PA. 


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-value is very very low, indicating that the results are statistically significant. The adjusted r-squared is 67%. The coefficient is negative. All three of them together indicate that there is a significant relationship with higher FARMS % indicating lower math scores. I think this model is worth using because of the very low p-value and higher percentage for adjusted r-squared. The PA model was not as sure of the relationship between the two variables and the results were not as clear as in Moco MD. 


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

Fox Chapel Elementary -- 30 residual
Almost equally split in population between Asian, Black, White and Hispanic
10% of people below the poverty line
Higher median household income that the state average
I don't see anything specific that explains why the school is over-perfoming the model.
I would ask the principal about their preparation for the math test and if that might be different than other schools in Montgomery County.

Somerset Elementary -- -26 residual 
70% white, very small percentage Black, Hispanic, Asian
4.6% of people below the poverty line
Almost double the MD median household income
There is a high percentage of people that are from out of the country compared to the state and national average so the only thing I can think of is students not doing well because English isn't their first language. 
I would want to ask the principal about the percent of students that English isn't their first language and also what preparation goes into getting the students ready for the test. 

### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The claim that elementary schools with fewer students on FARMS do better on standardized math tests is supported more by the Maryland data than Pennsylvania, but the amount of high residuals in the Maryland data makes me weary to confirm the board member's claim. I would want to visit the schools with the highest positive and negative residuals, and maybe the second highest ones as well because they deviate the most from the model and might prove the model wrong. If I called the school board member for comment, I would ask if there is a standardized preparation model in the schools for the math tests or if the individual schools are allowed to choose how they want to prepare the students. I would also ask if there are different leveled classes for math and if more students on FARMS ended up in lower level classes and therefore do worse on the standardized tests. 
