---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

From a reporting standpoint, I would say that the most important categories to focus on would be homicides and assault shooting/weapon. While there are many instances of accidents, auto stolen, and theft, since they are so common they would be less likely to be a news story. I would say there is a pretty wide range of numbers from month to month for robbery vehicle and B&E residental, which might make it more difficult to summarize. 

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov +Dec) |> mutate(average_incidents = total_incidents / 12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The categories that have the highest average number of incidents per month are accidents, accident impound, auto stolen, theft, and theft from auto. The categories that have the lowest average number of incidents per month are homicide, B&E other, and robbery residential since there is no average because of low numbers. I am wondering what the "other" of B&E entails, since the other two categories (commercial and residential) have much higher averages. I think it also worth taking note of the difference between auto stolen and auto stolen and recovered which is stark. 


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change = ((Jun-average_incidents)/average_incidents *100), (july_change = (Jul-average_incidents)/average_incidents *100), (august_change = (Aug-average_incidents)/average_incidents *100))
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

While some categories show an increase in crime, the majority of them show a decrease. This disproves the theory that crime is more likely to occur in the summer months, or at least invites in further investigation into specific crimes that might occur more in the summer. 

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

There were an average of 15 assault shootings each month this year in Prince George's County. 

July, August and October each saw 21 instances, with most other months seeing a few less. January had the lowest number of assault shootings at 5. There was a total of 188 assault shootings this year. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
