---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <-read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

What I immediately found interesting was that some people who work at the White House are marked "Detailee" instead of "Employee." I wonder what the distinction is because they seem to make a similar amount or more than employees. As for the salaries, there seem to be some inconsistencies because some of them say 0 while another person in the same position has a salary listed. If these are a mistake, it would cause an issue with analyzing the data because figures like mean and median will be skewed. From a reporting standpoint, I think it is notable that some "assistants" are making much more money than directors and other positions. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
install.packages("tidyverse")
library(tidyverse)
wh_salaries |> summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
min_salary = min(SALARY),
max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median are relatively close together, which confirms that the mean is a pretty good representation of the data despite there being outliers. The minumum salary of 0 is an outlier as it brings down the mean and median and creates a much bigger range than if it was not in there with the rest of the data. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I would take the route of the White House potentially omitting salaries from their reports because of the multiple instances of 0 being listed as a salary. The next step I would take is to analyze the positions of the people that are listed as 0 and see if there is any relation between them.
