---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
set.seed(42)

sample_5 <- mandms |> sample_n(5)
```


### Task 4: Calculate the mean and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
sample_5 |> summarize(
  mean_5 = mean(red),
  sd_5 = sd(red)
)
```


### Task 5: Calculate the mean and standard deviation of the population, using the same measure as in Task 4

```{r}
mandms |> summarize(
   mean_pop = mean(red),
  sd_pop = sd(red)
)
```


### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The two sets of means and standard deviations are pretty far apart. The sample has a mean that is higher than the mean of the population and a standard deviation that is much lower than the standard deviation of the population. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This tells me that there is some level of variance in the data, and since we took such a small sample, it wasn't representative of the entire dataset. Random sampling is more effective when you take a larger sample or try the same sample size multiple times to get a better representation of the data. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
