---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Amanda Sinofsky"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
health_scores <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- health_scores |> summarize (mean_comp = mean(compliance_score), sd_comp = sd(compliance_score), min_comp = min(compliance_score),
max_comp = max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
health_scores |>
ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 6) +
geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
labs(
    title = "Distribution of Compliance Scores for Restaurants",
    subtitle = "Red line = mean",
    x = "Compliance Score",
    y = "Count",
  )
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**

The histogram tells us that the distribution is not normal and the majority of the data clusters to the right, making the distribution skewed to the left. There is a standard deviation of about 5.8, but since this is not a normal distribution we can see that the majority of the data falls to the right of the mean. There is a dip between 90 and 100 in scores which I think is worth taking a look at. I also would like to know if there is a certain type of restaurant or a certain location in which restaurants get higher compliance scores. 
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> summarize(
  mean_bus = mean(bus),
  mean_rail = mean(rail),
  sd_bus = sd(bus),
  sd_rail = sd(rail)
)
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily |> mutate(
  mean_bus = mean(bus),
  mean_rail = mean(rail),
  sd_bus = sd(bus),
  sd_rail = sd(rail),
  rail_zscore = (total - mean_rail) / sd_rail,
  bus_zscore = (total - mean_bus) / sd_bus
)
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_unusual <- wmata_zscores |>
  filter(rail_zscore > 2)
```

There are 301 unusual days if we classify unusual as an absolute value greater than 2, and these are all positive z-scores. The most extreme is on October 29, 2024 with a z-score of 6.1.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)

sample_91 <- wmata_daily |> sample_n(91)

sample_91 |> summarize(
  mean_91_bus = mean(bus),
  mean_91_rail = mean(rail),
  sd_91_bus = sd(bus),
  sd_91_rail = sd(rail))
```
The sample means and standard deviations are slightly higher than those of the actual population. This suggests that random sampling can be a good estimate of the makeup of a population, but a bigger sample will give a more accurate representation of a population. Since our sample of 91 is only 25% of the population, it is not as accurate than if we used a bigger sample. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

On most of the days with unusual rail z-scores, there is an unusually high bus z-score as well. A lot of the dates of the most unusual z-scores are also in October. This could indicate that the weather was particularly bad that month or there was something else going on in October that contributed to more people using public transportation than normal. 
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |> mutate(
  theft_rate = (thefts_2023 / population) *10000

)
```
Each county has a different population, so getting a rate of how many thefts occur per person in that county is more accurate than simply counting the number of thefts in each county. Our results would not be as compelling if we did that because every county's population is different and cannot be compared accurately. 

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts <- md_car_thefts |> mutate(
  median_theft = median(theft_rate),
  total_thefts = sum(thefts_2023)
)

 md_rates_above <- md_car_thefts |>
  filter(theft_rate > median_theft)

```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

The distribution of car thefts in Maryland is almost normal except for Baltimore City. These calculations reveal that Baltimore City has a theft rate of 123 per 10,000 people, while the county right below that is at 89 in Prince George's County. 

Lede: Baltimore City has highest car theft rate out of Maryland counties at 123 per 10,000 residents.

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points) 

I would prioritize using the empirical method to see how many visitors are encompassed within 68, 95, and 99.7% of the visitors per day. This is a pretty high standard deviation that shows there is a wide range of variance in how many visitors are there. I would want to see if there were any unusual days and I also would want the average visitors of each individual swimming pool in the community. I would also want a complete count rather than a random sample of 60 days. 


### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would use z-scores to see if any are unusually low in certain pools. This is more useful than ranking by attendance because it standardizes the results regardless of the amount of visitors. Some pools might have more people that live near them which would be unfair to compare to smaller pools or pools that serve a smaller community. Z-scores allow for accurate comparison between each pool without simply counting visitors. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would ask if they used the same 60 days for each pool, and how many days of the summer the pool is open compared to the 60 days that they sampled from. I also would consider if any of the 60 days were holidays like Memorial Day, Fourth of July, or Labor Day that could skew the results. In my story, I would raise the concern of each pool having a different number of typical patrons or the size of the pool/how many people it could accommodate, and the part of the town that the pool was in and how many people live in that area. 

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
